---
title: "Regression Analysis"
output:   
  html_document:
    toc: true
    toc_float: true
---

```{r,include=FALSE,message=FALSE,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(broom)
library(plotly)
library(highcharter)
library(rworldmap)
library(gganimate)
library(transformr)
library(corrplot)
library(leaps)
library(kableExtra)
library(modelr)
library(mgcv)
knitr::opts_chunk$set(
 echo = FALSE,
 fig.width = 7, 
 fig.height = 5,
 fig.asp = 0.6,
 out.width = "60%")
theme_set(theme_bw() + 
          theme(legend.position = "bottom",
                legend.title = element_blank(),
                plot.title = element_text(hjust = 0.5, size = 15),
                plot.subtitle = element_text(hjust = 0.5, size = 12)))


```

### Data Import and Cleaning


#### HIV Deaths By Country Dataset

```{r, message = FALSE,warning = FALSE}
deaths = read_csv("data/final_hiv_deaths.csv") %>%
 janitor::clean_names() %>%
 select("location_id", "location_name", starts_with("sex"), starts_with("age"), "year", "val", "upper", "lower") %>%
 rename(country_name = location_name) %>% 
  mutate(year = as.character(year))
deaths %>%
 summarize(n_countries = n_distinct(country_name),
           n_years = n_distinct(year),
           col_name = colnames(deaths),
           n_missing = map(deaths, ~sum(is.na(.)))) %>%
 unnest(n_missing)
```

The death dataset contains HIV death count data from 204 distinct countries, in the 30 years between 1990 and 2019.

The deaths dataset contains `r nrow(deaths)` rows and `r ncol(deaths)` columns, with each row representing a single record of number of HIV deaths in a country stratified by year, sex and age group. Variables in the deaths dataset are as follows: `r colnames(deaths)`. 


#### GDP Per Capita By Country Dataset

```{r, message = FALSE,warning = FALSE}
gdp = read_csv("data/final_gdp_per_capita.csv", skip = 4, col_names = TRUE) %>% 
  janitor::clean_names() %>% 
  drop_na(location_id) 
names(gdp) <- sub('^x', '', names(gdp))
gdp_clean = gdp %>%  
  select("location_id", "country_name", "1990":"2019") %>% 
  pivot_longer(cols = "1990":"2019",
               names_to = "year",
               values_to = "GDP_per_capita") 
```
The raw GDP dataset contains GDP per capita (US dollars) from 200 distinct countries, from 1960 to 2021. 

The dataset contains `r nrow(gdp)` rows and `r ncol(gdp)` columns, with each row representing a single record of a country's GDP per capital (US dollars) in a given year. 


#### Population By Country Dataset
```{r, message = FALSE,warning = FALSE}
population = read_csv("data/final_population_total.csv", skip = 4, col_names = TRUE)  %>% 
  janitor::clean_names() %>% 
  drop_na(location_id) 
names(population) <- sub('^x', '', names(population))
population_clean = population %>%  
  select("location_id", "1990":"2019") %>% 
  pivot_longer(cols = "1990":"2019",
               names_to = "year",
               values_to = "Population") 
```

The raw population dataset contains total population from 200 distinct countries, from 1960 to 2021. 

The dataset contains `r nrow(population)` rows and `r ncol(population)` columns, with each row representing a single record of a country's population in a given year. 



### Merging 3 datasets to produce the "maindata" dataset for visualization and analysis

First, we merge the GDP Per Capita By Country dataset (`gdp_clean`) and Population By Country Dataset (`population_clean`) into a dataset called `gdp_population`. This combined dataset contains data from 200 distinct countries between 1990 and 2019.

```{r, message = FALSE,warning = FALSE}
gdp_population = inner_join(gdp_clean, population_clean, by = c("location_id", "year")) %>%
 janitor::clean_names() 
gdp_population %>%
 summarize(n_countries = n_distinct(country_name),
           n_years = n_distinct(year),
           col_name = colnames(gdp_population),
           n_missing = map(gdp_population, ~sum(is.na(.)))) %>%
 unnest(n_missing)
```

Next, we merge the `gdp_population` dataset with the HIV Deaths dataset (`deaths`) into a `maindata` dataset by matching on the countries and years and using `left_join`. We save this as a CSV for further analysis.

```{r, message = FALSE,warning = FALSE}
maindata = left_join(gdp_population, deaths, by = c("year","location_id")) %>% 
  select(-country_name.x, -age_id) %>% 
  rename(country_name = country_name.y) %>% 
  select(location_id, country_name, everything())
maindata %>% 
  summarize(n_countries = n_distinct(location_id),
            n_years = n_distinct(year),
            col_name = colnames(maindata),
            n_missing = map(maindata, ~sum(is.na(.)))) %>% 
  unnest(n_missing)
write_csv(maindata, 'data/maindata.csv')
```

`maindata` has `r nrow(maindata)` observations and `r ncol(maindata)` features. Basic features we are interested in include:

- **location_id**: id corresponding to a country_name
- **sex_id**: id corresponding to sex
- **sex_name**: Male or Female
- **age_name**: Aae groups including "0-9 years", "10-24 years",  "25-49 years", "50-74 years", and "75+ years".
- **year**: Year of the records
- **val**: Point estimate of number of HIV/AIDS deaths
- **upper**: Upper estimate of number of HIV/AIDS deaths
- **lower**: Lower estimate of number of HIV/AIDS deaths
- **gdp_per_capita** GDP per capita in 2022 US dollars
- **population** population total
- **country_name** name of country


### Statistical analysis

#### Create "motality_per_100k" and "log_motality_per_100k" variables

```{r}
regression = maindata %>% 
  mutate(motality_per_100k = val / (population/100000),
         log_motality_per_100k = log(motality_per_100k),
         age_name = as.factor(age_name),
         sex_name = as.factor(sex_name),
         year = as.numeric(year),
         year_2004 = if_else(year <= 2004, 1, 0),
         year_2004 = as.factor(year_2004),
         country_name = as.factor(country_name)) 

regression %>% 
  summarize(col_name = colnames(regression),
            n_missing = map(regression, ~sum(is.na(.)))) %>% 
  unnest(n_missing)

```
### Plots of the distribution to main outcome of interest(HIV mortality per 100k population).
```{r}
regression %>% 
  ggplot(aes(x = motality_per_100k)) +
  geom_histogram(color = "darkblue", fill = "lightblue") +
  ggtitle("Histogram of HIV deaths per 100k") +
  geom_vline( aes(xintercept = mean(motality_per_100k)),
             linetype = "dashed") +
  ylab("") +
  xlab("Estimated HIV deaths per 100k")


regression %>% 
  ggplot(aes(x = log_motality_per_100k)) +
  geom_histogram(color = "darkblue", fill = "lightblue") +
  ggtitle("Histogram of log(HIV deaths per 100k)") +
  geom_vline(aes(xintercept = mean(log_motality_per_100k)),
             linetype = "dashed") +
  ylab("") +
  xlab("Estimated log(HIV deaths per 100k)")
```
### Logistic regression models

### model1:

$$ log(motality/100000) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year + \beta_4 gdp $$
```{r}
regression_logit = regression %>%
  mutate(log_motality_per_100k = if_else(log_motality_per_100k == -Inf, 0.01, log_motality_per_100k))
  
model1_logit = lm(log_motality_per_100k ~ age_name + sex_name + year + gdp_per_capita, data = regression_logit) 

kable(summary(model1_logit)$coefficients)

```

### model2:

$$ log(motality/100000) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year + \beta_4year_{2004} + \beta_5gdp + \beta_6 year*year_{2004}  $$
```{r}
  
model2_logit = lm(log_motality_per_100k ~ age_name + sex_name + year + year_2004 + gdp_per_capita + year*year_2004, data = regression_logit)

kable(summary(model2_logit)$coefficients)

```

### model3:

$$ log(motality/100000) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year_{2004} + \beta_4 gdp + \beta_5 age*sex $$

```{r}
model3_logit = lm(log_motality_per_100k ~ age_name + sex_name + year_2004 + gdp_per_capita + age_name*sex_name, data = regression_logit)

kable(summary(model3_logit)$coefficients)

```

### model4:
$$ log(motality/100000) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year_{2004} + \beta_4 gdp + \beta_5 age*sex + \beta_6 country$$

```{r}
model4_logit = lm(log_motality_per_100k ~ age_name + sex_name + year_2004 + gdp_per_capita + age_name*sex_name + country_name, data = regression_logit)

kable(summary(model4_logit)$coefficients)

```

#### Poisson Regression models:

### model1:

$$ log(E(Y_i|age_i,sex_i,year_i,gdp_i)) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year + \beta_4 gdp $$
```{r}
regression_poisson = regression %>% 
  mutate(val = as.integer(val))

model1_poi = glm(val ~ age_name + sex_name + year + gdp_per_capita, family = "poisson", data = regression_poisson)

kable(summary(model1_poi)$coefficients)
```
### model2:

$$ log(E(Y_i|age_i,sex_i,year_i,year_{2004i},gdp_i)) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year + \beta_4 year_{2004} + \beta_5 gdp + \beta_6 year *year_{2004}$$
```{r}
model2_poi = glm(val ~ age_name + sex_name + year + year_2004 + gdp_per_capita + year*year_2004, family = "poisson", data = regression_poisson)

kable(summary(model2_poi)$coefficients)
```

### model3:

$$ log(E(Y_i|age_i,sex_i,year_{2004i},gdp_i)) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year_{2004} + \beta_4 gdp + \beta_5 age *sex$$
```{r}
model3_poi = glm(val ~ age_name + sex_name + year_2004 + gdp_per_capita + age_name*sex_name, data = regression_poisson)

kable(summary(model3_logit)$coefficients)
```

### model4:

$$ log(E(Y_i|age_i,sex_i,year_{2004i},gdp_i,country_i)) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year_{2004} + \beta_4 gdp + \beta_5 age *sex + \beta_6 country$$
```{r}
model4_logit = glm(val ~ age_name + sex_name + year_2004 + gdp_per_capita + age_name*sex_name + country_name, data = regression_poisson)

kable(summary(model4_logit)$coefficients)
```

```{r}

```

