---
title: "Regression Analysis"
output:   
  html_document:
    toc: true
    toc_float: true
---

```{r,include=FALSE,message=FALSE,echo=FALSE}
library(tidyverse)
library(ggplot2)
library(broom)
library(plotly)
library(highcharter)
library(rworldmap)
library(gganimate)
library(transformr)
library(corrplot)
library(leaps)
library(kableExtra)
library(modelr)
library(mgcv)
knitr::opts_chunk$set(
 echo = FALSE,
 fig.width = 7, 
 fig.height = 5,
 fig.asp = 0.6,
 out.width = "60%")
theme_set(theme_bw() + 
          theme(legend.position = "bottom",
                legend.title = element_blank(),
                plot.title = element_text(hjust = 0.5, size = 15),
                plot.subtitle = element_text(hjust = 0.5, size = 12)))


```

### Data Import and Cleaning

#### HIV Deaths By Country Dataset

```{r, message = FALSE,warning = FALSE}
deaths = read_csv("data/final_hiv_deaths.csv") %>%
 janitor::clean_names() %>%
 select("location_id", "location_name", starts_with("sex"), starts_with("age"), "year", "val", "upper", "lower") %>%
 rename(country_name = location_name) %>% 
  mutate(year = as.character(year))
deaths %>%
 summarize(n_countries = n_distinct(country_name),
           n_years = n_distinct(year),
           col_name = colnames(deaths),
           n_missing = map(deaths, ~sum(is.na(.)))) %>%
 unnest(n_missing)
```

The death dataset contains HIV death count data from 204 distinct countries, in the 30 years between 1990 and 2019.

The deaths dataset contains `r nrow(deaths)` rows and `r ncol(deaths)` columns, with each row representing a single record of number of HIV deaths in a country stratified by year, sex and age group. Variables in the deaths dataset are as follows: `r colnames(deaths)`. 


#### GDP Per Capita By Country Dataset

```{r, message = FALSE,warning = FALSE}
gdp = read_csv("data/final_gdp_per_capita.csv", skip = 4, col_names = TRUE) %>% 
  janitor::clean_names() %>% 
  drop_na(location_id) 
names(gdp) <- sub('^x', '', names(gdp))
gdp_clean = gdp %>%  
  select("location_id", "country_name", "1990":"2019") %>% 
  pivot_longer(cols = "1990":"2019",
               names_to = "year",
               values_to = "GDP_per_capita") 
```
The raw GDP dataset contains GDP per capita (US dollars) from 200 distinct countries, from 1960 to 2021. 

The dataset contains `r nrow(gdp)` rows and `r ncol(gdp)` columns, with each row representing a single record of a country's GDP per capital (US dollars) in a given year. 


#### Population By Country Dataset
```{r, message = FALSE,warning = FALSE}
population = read_csv("data/final_population_total.csv", skip = 4, col_names = TRUE)  %>% 
  janitor::clean_names() %>% 
  drop_na(location_id) 
names(population) <- sub('^x', '', names(population))
population_clean = population %>%  
  select("location_id", "1990":"2019") %>% 
  pivot_longer(cols = "1990":"2019",
               names_to = "year",
               values_to = "Population") 
```

The raw population dataset contains total population from 200 distinct countries, from 1960 to 2021. 

The dataset contains `r nrow(population)` rows and `r ncol(population)` columns, with each row representing a single record of a country's population in a given year. 


### Merging 3 datasets to produce the "maindata" dataset for visualization and analysis

First, we merge the GDP Per Capita By Country dataset (`gdp_clean`) and Population By Country Dataset (`population_clean`) into a dataset called `gdp_population`. This combined dataset contains data from 200 distinct countries between 1990 and 2019.

```{r, message = FALSE,warning = FALSE}
gdp_population = inner_join(gdp_clean, population_clean, by = c("location_id", "year")) %>%
 janitor::clean_names() 
gdp_population %>%
 summarize(n_countries = n_distinct(country_name),
           n_years = n_distinct(year),
           col_name = colnames(gdp_population),
           n_missing = map(gdp_population, ~sum(is.na(.)))) %>%
 unnest(n_missing)
```

Next, we merge the `gdp_population` dataset with the HIV Deaths dataset (`deaths`) into a `maindata` dataset by matching on the countries and years and using `left_join`. We save this as a CSV for further analysis.

```{r, message = FALSE,warning = FALSE}
maindata = left_join(gdp_population, deaths, by = c("year","location_id")) %>% 
  select(-country_name.x, -age_id) %>% 
  rename(country_name = country_name.y) %>% 
  select(location_id, country_name, everything())
maindata %>% 
  summarize(n_countries = n_distinct(location_id),
            n_years = n_distinct(year),
            col_name = colnames(maindata),
            n_missing = map(maindata, ~sum(is.na(.)))) %>% 
  unnest(n_missing)
write_csv(maindata, 'data/maindata.csv')
```

`maindata` has `r nrow(maindata)` observations and `r ncol(maindata)` features. Basic features we are interested in include:

- **location_id**: id corresponding to a country_name
- **sex_id**: id corresponding to sex
- **sex_name**: Male or Female
- **age_name**: Aae groups including "0-9 years", "10-24 years",  "25-49 years", "50-74 years", and "75+ years".
- **year**: Year of the records
- **val**: Point estimate of number of HIV/AIDS deaths
- **upper**: Upper estimate of number of HIV/AIDS deaths
- **lower**: Lower estimate of number of HIV/AIDS deaths
- **gdp_per_capita** GDP per capita in 2022 US dollars
- **population** population total
- **country_name** name of country


### Statistical analysis

#### Cleaning and tidying data for regression analysis

```{r}
regression = maindata %>% 
  mutate(mortality_per_100k = val / (population/100000),
         log_mortality_per_100k = log(mortality_per_100k),
         age_name = as.factor(age_name),
         sex_name = as.factor(sex_name),
         year = as.numeric(year),
         year_2004 = if_else(year <= 2004, 1, 0),
         year_2004 = as.factor(year_2004),
         country_name = as.factor(country_name)) 

regression %>% 
  summarize(col_name = colnames(regression),
            n_missing = map(regression, ~sum(is.na(.)))) %>% 
  unnest(n_missing)

```

We created "mortality_per_100k" (mortality per 100k population), and "log_mortality_per_100k" (log of mortality per 100k population) variables to explore the distribution of the outcome of interest: HIV mortality. We also created the "year_2004" variable to explore the change of HIV mortality before and after 2004.

### Plots of the distribution to main outcome of interest(HIV mortality per 100k population).
```{r}
regression %>% 
  ggplot(aes(x = mortality_per_100k)) +
  geom_histogram(color = "darkblue", fill = "lightblue") +
  ggtitle("Histogram of HIV deaths per 100k") +
  geom_vline( aes(xintercept = mean(mortality_per_100k)),
             linetype = "dashed") +
  ylab("") +
  xlab("Estimated HIV deaths per 100k")


regression %>% 
  ggplot(aes(x = log_mortality_per_100k)) +
  geom_histogram(color = "darkblue", fill = "lightblue") +
  ggtitle("Histogram of log(HIV deaths per 100k)") +
  geom_vline(aes(xintercept = mean(log_mortality_per_100k)),
             linetype = "dashed") +
  ylab("") +
  xlab("Estimated log(HIV deaths per 100k)")
```
The above graphs show that our outcome variable (the HIV deaths per 100k) is extremely right skewed, but after log transformation, it becomes somewhat normally distributed so we could run linear models with the log transformed outcome.


### Linear regression models

### Hypothesis 1:
There is statistically significant relationship between the HIV deaths per 100k and age groups, sex, year, and the country's gdp respectively. 

### model 1:

$$ log(motality/100000) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year + \beta_4 gdp $$
```{r}
regression_logit = regression %>%
  mutate(log_mortality_per_100k = if_else(log_mortality_per_100k == -Inf, 0.01, log_mortality_per_100k))
  
model1_logit = lm(log_mortality_per_100k ~ age_name + sex_name + year + gdp_per_capita, data = regression_logit) %>% 
  broom::tidy() %>% 
  mutate(exp_est = exp(estimate))

knitr::kable(model1_logit)

```

Interpretation:
The 10-24 years old has 0.83 times the HIV mortality per 100k population compared to the 0-9 years old holding sex, year and gdp in constant.
The 25-49 years old has 9.26 times the HIV mortality per 100k population compared to the 0-9 age years old holding sex, year and gdp in constant.
The 50-74 years old has 2.70 times the HIV mortality per 100k population compared to the 0-9 age years old holding sex, year and gdp in constant.
The 75+ years old has 0.06 times the HIV mortality per 100k population compared to the 0-9 age years old holding sex, year and gdp in constant.
The HIV mortality per 100k population will increase by 4% with 1 year increase between 1990-2019 holding age, sex, and gdp in constant.
The the HIV mortality per 100k population will decrease by 0.004% with 1 unit increase of gdp holding age, sex, and year in constant.

The associations between the HIV deaths per 100k and age groups, sex, year, and the country's gdp respectively are all statistically significant at the level of 5%.


### Hypothesis 2:
There is a significant change of the HIV deaths per 100k before and after the year 2004 controlling for age groups, sex, and gdp.

### model 2:
$$ log(motality/100000) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year_{2004} + \beta_4gdp  $$
```{r}
  
model2_logit = lm(log_mortality_per_100k ~ age_name + sex_name + year_2004 + gdp_per_capita, data = regression_logit) %>% 
  broom::tidy() %>% 
  mutate(exp_est = exp(estimate))

knitr::kable(model2_logit)

```
Interpretation: 
The HIV deaths per 100k is 44.51% lower before 2004 than after 2004 after controlling for age, sex, and gdp at the level of 5% significance. 

### Hypothesis 3:
There is an interaction between age and sex. 

### model 3:

$$ log(motality/100000) = \beta_0 + \beta_1 age + \beta_2 sex + \beta_3 year + \beta_4 gdp + \beta_5 age*sex $$

```{r}
model1_compare = lm(log_mortality_per_100k ~ age_name + sex_name + year + gdp_per_capita, data = regression_logit)

model3_logit = lm(log_mortality_per_100k ~ age_name + sex_name + year + gdp_per_capita + age_name*sex_name, data = regression_logit) 

anova(model1_compare, model3_logit) 

```

Interpretation:
There is an interaction between age groups and sex at the 5% significance level, which means that gender can have different impacts on the HIV deaths per 100k regarding to different age groups.


